Article Title

### 2015-02-24 manuscript.tex

Pearson correlations between the spiking activity of pairs of neurons are among the most familiar measures of functional connectivity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  In particular, \emph{noise correlations}, \emph{i.e.}\;the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Nirenberg:2003, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect aspects of synaptic connectivity \cite{Gerstein:1964}.  Interest in neural correlations has been sustained by a series of discoveries of their nontrivial relationships to various aspects of circuit organization such as the physical distances between the neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010, Rothschild:2010, Ko:2011, Smith:2013b}, cell types \cite{Hofer:2011}, cortical layer specificity \cite{Hansen:2012, Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011, Ko:2013}, changes due to sensory stimulation and global brain states \cite{Greenberg:2008, Goard:2009, Kohn:2009, Rothschild:2010, Ecker:2014, Renart:2010}.

@@

Improved estimation of the correlation matrix is beneficial in itself. For example, improved estimates can be used to optimize  decoding of the population activity \cite{Friedman:1989, Berens:2012}. But reduced estimation error is not the only benefit of regularization.  Finding the most efficient among many regularized estimators leads to insights about the system itself: the structure of the most efficient estimator is a parsimonious representation of the regularities in the data.

@@

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is a $p\times 1$ vector of firing rates in time bin $t$, and an unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix} is
\begin{equation}\label{eq:sample}
    C_{\sf sample} = \frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
    \end{equation}
where $n$ is the number of time bins in $T$. The sample covariance matrix yields an unbiased estimate of the covariance matrix ($\E{C_{\sf sample}}=\Sigma$) if the mean, $\bar x$, is independent of the data sample, $\{x(t): t\in T$\}. When the mean is estimated from the same sample, $C_{\sf sample}$ becomes biased toward zero.  However, in all the cases where unbiasedness is required, we will estimate the mean from a separate sample.

@@

Ising models have been used to infer functional connectivity from neuronal spike trains \cite{Hertz:2011}. 
Conveniently, the Ising model has equivalent mathematical form to the Gaussian distribution,
\begin{equation}\label{eq:ising}
    x\sim \frac 1 {Z(J,h)} \exp\left(\frac 1 2 x^\T J x + h^\T x \right)
\end{equation}
but the Ising model is defined on the multivariate binary domain rather than the continuous domain. 
Both models are maximum-entropy models constrained to match the mean firing rates and the covariance matrix \cite{Jaynes:1957}.
The partition function $Z(J,h)$ normalizes the distribution on the models' respective domains. 
In the Gaussian model, the matrix $-J^{-1}$ is the covariance matrix; and the mean values are $\mu=J^{-1}h$.  
For the Ising model, $J$ is the matrix of pairwise interactions and $h$ is the vector of the cells' individual activity drives, although they do not have a simple relationship to the means and the covariance matrix. 
Both distributions have the same structure of pairwise conditional dependencies: zeros in the matrix $J$ indicate conditional independence between the corresponding pair of neurons.

Indeed, despite their considerable departure from strictly linear conditional dependencies, Ising models yielded the same relationships between the performances of the covariance estimators as the Gaussian models in cross-validation (Fig.~\ref{fig:2}). Identical interaction matrices $J$  of the joint distributions over the observable and latent variables were used for both the Gaussian and the Ising models.

@@

\subsection*{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of lightly sedated mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (Fig.~\ref{fig:3} A--B) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (Fig.~\ref{fig:3} C and D).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.~\ref{fig:3} C). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (Fig.~\ref{fig:3} D). To estimate the noise correlation matrix, we presented only two distinct directions in some experiments or five directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the correlation matrix.   The sample correlation coefficients were largely positive and low (Fig.~\ref{fig:3} E and F). The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean across sites of 0.018 (Fig.~\ref{fig:6} D).

@@

Inference of the conditional dependencies also depends on the completeness of the recorded population:  To equate conditional dependency to direct interaction between two neurons, we must record from all neurons with which the pair interacts. Unobserved portions of the circuit may manifest as conditional dependencies between observed neurons that do not interact. For this reason, statistical models of population activity have been most successfully applied to \emph{in vitro} preparations of the retina or cell cultures where high-quality recordings from the complete populations were available \cite{Pillow:2008}. In cortical tissue, electrode arrays record from a small fraction of cells in a given volume, limiting the validity of inference of the pairwise conditional dependencies. Perhaps for this reason, partial correlations have not, until now, been used to describe the functional connectivity in cortical populations.

Two-photon imaging of population calcium signals presents unique advantages for the estimation of functional connectivity.  While the temporal resolution of calcium signals is limited by calcium dye kinetics, fast imaging techniques combined with spike inference algorithms provide millisecond-scale temporal resolution of single action potentials \cite{Grewe:2010}. However, such high temporal precision comes at the cost of the accuracy of inferred spike rates.  Better accuracy is achieved when calcium signals are analyzed on scales of tens of milliseconds \cite{Cotton:2013, Theis:2014}.  The major advantage of calcium imaging is its ability to characterize the spatial arrangement and types of recorded cells.  Recently, advanced imaging techniques have allowed recording from nearly every cell in a volume of cortical tissue  \emph{in vivo} \cite{Katona:2012, Cotton:2013} and even from entire nervous systems \cite{Leung:2013, Ahrens:2013}.  These techniques may provide more incisive measurements of functional connectivity than electrophysiological recordings.

@@

\subsection*{Functional connectivity as coactivations}
Another approach to describing the functional connectivity of a circuit is to isolate individual patterns of multineuronal coactivations \cite{Gerstein:1989, Chapin:1999, Peyrache:2010, Ch:2010, Lopes:2011, Lopes:2013}. Depending on the method of their extraction, coactivation patterns may be referred to as \emph{assemblies}, \emph{factor loadings}, \emph{principal components}, \emph{independent components}, \emph{activity modes}, \emph{eigenvectors}, or \emph{coactivation maps}. Coactivation patterns could be interpreted as signatures of Hebbian cell assemblies \cite{Gerstein:1989, Ch:2010}, \emph{i.e.}\;groups of tightly interconnected groups of cells involved in a common computation.  Coactivation patterns could also result from shared input from unobserved parts of the circuit, or global network fluctuations modulating the activity of the local circuit \cite{Okun:2012, Ecker:2014}.

@@

\subsection*{Combining pairwise interactions and coactivations}
In the effort to account for the joint activity patterns that are poorly explained by pairwise interactions, investigators have augmented models of pairwise interactions with additional mechanisms such as latent variables, higher-order correlations, or global network fluctuations \cite{Ganmor:2011, Tkacik:2013, Pfau:2013, Koster:2013, Ecker:2014}.

@@

We showed that among several models a sparse network of linear interactions with several latent inputs yielded the best estimates of the noise covariance matrix for cortical microcircuits.  This finding is valuable in itself: improved estimates of the noise covariance matrix for large datasets are important in order to understand the role of noise correlations in population coding \cite{Abbott:1999, Sompolinsky:2001, Averbeck:2006, Josic:2009, Ecker:2011}

@@

\subsection*{Surgery and two-photon imaging}
The surgical procedures and data acquisition were performed as described in \cite{Cotton:2013}: C57BL/6J mice (aged p40--60) were used. For surgery, animals were initially anesthetized with isoflurane (3\%). During the experiments, animals were sedated with a mixture of fentanyl (0.05 mg/kg), midazolam (5 mg/kg), and medetomidine (0.5 mg/kg), with boosts of half the initial dose every 3 hours.  A craniotomy was performed over the right primary visual cortex.  Membrane-permeant calcium indicator Oregon Green 488 BAPTA-1 AM (OGB-1, Invitrogen) was loaded by bolus injection.  The craniotomy was sealed using a glass coverslip secured with dental cement.

@@

\subsection*{Visual stimulus}
The visual stimulus consisted of full-field drifting gratings with 90\% contrast, luminance of 10 cd/m$^2$, spatial frequency of 0.08 cycles/degree, and temporal frequency of 2 cycles/s. Two types of stimuli were presented for each imaging site: First, directional tuning was mapped using a pseudo-random sequence of drifting gratings at sixteen directions of motion, 500 ms per direction, without blanks, with 12--30 trials for each direction of motion.  Second, to measure correlations, the stimulus was modified to include only two directions of motion (in 9 datasets) or five directions (in 22 datasets) and the gratings were presented for 1 second and were separated by 1-second blanks, with 100--300 trials for each direction of motion.

@@

In estimator $C_{\sf factor}$ (Eq.~\ref{eq:c-factor}), the low-rank matrix $L$ and the diagonal matrix $D$ are found by solving the minimization problem
\begin{equation}
(L,D) = \argmin\limits_{\hat L,\hat D} \loss{\hat L + \hat D,C_{\sf sample}},
\end{equation}
by an expectation-maximization (EM) algorithm with specified rank of $L$. In addition, the diagonal matrix of individual variances is shrunk toward its mean value according to Eq.~\ref{eq:c-factor}.

>>>>

Pearson correlations between the spiking activity of pairs of neurons are among the most familiar measures of functional connectivity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  In particular, \emph{noise correlations}, \emph{i.e.}\;the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Nirenberg:2003, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect aspects of synaptic connectivity \cite{Gerstein:1964}.  Interest in neural correlations has been sustained by a series of discoveries of their nontrivial relationships to various aspects of circuit organization such as the physical distances between the neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010, Rothschild:2010, Ko:2011, Smith:2013b}, cell types \cite{Hofer:2011}, cortical layer specificity \cite{Hansen:2012, Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011, Ko:2013}, changes due to sensory stimulation and global brain states \cite{Greenberg:2008, Goard:2009, Kohn:2009, Rothschild:2010, Ecker:2010, Renart:2010}.

@@

Improved estimation of the correlation matrix is beneficial in itself. For example, improved estimates can be used to optimize  decoding of the population activity \cite{Friedman:1989}. But reduced estimation error is not the only benefit of regularization.  Finding the most efficient among many regularized estimators leads to insights about the system itself: the structure of the most efficient estimator is a parsimonious representation of the regularities in the data.

@@

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is a $p\times 1$ vector of firing rates in time bin $t$, and an independent unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix},
\begin{equation}\label{eq:sample}
    C_{\sf sample} = \frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
    \end{equation}
where $n$ is the number of time bins in $T$, yields an unbiased estimate of the covariance matrix so that $\E{C_{\sf sample}}=\Sigma$.

When the mean is estimated from the same sample, $C_{\sf sample}$ becomes biased toward zero.  However, in all the cases when the unbiasedness is required in our study, the mean will be estimated from a separate sample.

@@

Ising models have been used to infer functional connectivity from neuronal spike trains \cite{Hertz:2011}. 
Conveniently, the Ising model is described by a probability distribution equivalent in form to the Gaussian distribution,
\begin{equation}\label{eq:ising}
    x\sim \frac 1 {Z(J,h)} \exp\left(\frac 1 2 x^\T J x + h^\T x \right).
\end{equation}
However, the Ising model is defined on the multivariate binary domain rather than the continuous domain. 
Both models are maximum-entropy models constrained to match mean firing rates and the covariance matrix \cite{Jaynes:1957}.
The partition function $Z(J,h)$ normalizes the distribution on the models' respective domains. 
In the Gaussian model, the matrix $-J^{-1}$ is the covariance matrix, and the mean values are $\mu=J^{-1}h$.  
For the Ising model, $J$ is the matrix of pairwise interactions, and $h$ is the vector of the cells' individual activity drives.  Neither $J$ nor $h$ have a simple relationship to the means and the covariance matrix. 
Both distributions have the same structure of pairwise conditional dependencies: zeros in the matrix $J$ indicate conditional independence between the corresponding pair of neurons.

Indeed, despite their considerable departure from strictly linear conditional dependencies, when we sampled data from an Ising models we observed similar relationships between the covariance estimators as in the Gaussian models in cross-validation (Fig.~\ref{fig:2}). Identical interaction matrices $J$  of the joint distributions over the observable and latent variables were used for both the Gaussian and the Ising models.

@@

\subsection*{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (Fig.~\ref{fig:3} A--B) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (Fig.~\ref{fig:3} C and D).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.~\ref{fig:3} C). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (Fig.~\ref{fig:3} D). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the correlation matrix.   The sample correlation coefficients were largely positive and low (Fig.~\ref{fig:3} E and F). The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean across sites of 0.018 (Fig.~\ref{fig:6} D).

@@

Inference of the conditional dependencies also depends on the completeness of the recorded population:  To equate conditional dependencies to direct interaction between neurons, we must record from all neurons with which the pair interacts. Unobserved portions of the circuit may manifest as conditional dependencies between observed neurons that do not interact. For this reason, statistical models of population activity have been most successfully applied to \emph{in vitro} preparations of the retina or cell cultures where high-quality recordings from the complete populations were available \cite{Pillow:2008}. In cortical tissue, electrode arrays record from a small fraction of cells in a given volume, limiting the validity of inference of the pairwise conditional dependencies. Perhaps for this reason, partial correlations have not, until now, been used to describe the functional connectivity in cortical populations.

Two-photon imaging of population calcium signals presents unique advantages for the estimation of functional connectivity.  While the temporal resolution of calcium signals is limited by calcium dye kinetics, fast imaging techniques combined with spike inference algorithms provide millisecond-scale temporal resolution of single action potentials \cite{Grewe:2010}. However, such high temporal precision comes at the cost of the accuracy of inferred spike rates.  Better accuracy is achieved when calcium signals are analyzed on scales of tens of milliseconds \cite{Cotton:2013}.  The major advantage of calcium imaging is its ability to characterize the spatial arrangement and types of recorded cells.  Recently, advanced imaging techniques have allowed recording from nearly every cell in a volume of cortical tissue  \emph{in vivo} \cite{Katona:2012, Cotton:2013} and even from entire nervous systems \cite{Leung:2013, Ahrens:2013}.  These techniques may provide more incisive measurements of functional connectivity than electrophysiological recordings.

@@

\subsection*{Functional connectivity as coactivations}
Another approach to describing the functional connectivity of a circuit is to isolate individual patterns of multineuronal coactivations \cite{Gerstein:1989, Chapin:1999, Peyrache:2010, Ch:2010, Lopes:2011, Lopes:2013}. Depending on the method of their extraction, coactivation patterns may be referred to as \emph{assemblies}, \emph{factor loadings}, \emph{principal components}, \emph{independent components}, \emph{activity modes}, \emph{eigenvectors}, or \emph{coactivation maps}. Coactivation patterns could be interpreted as signatures of Hebbian cell assemblies \cite{Gerstein:1989, Ch:2010}, \emph{i.e.}\;groups of tightly interconnected groups of cells involved in a common computation.  Coactivation patterns could also result from shared input from unobserved parts of the circuit, or global network fluctuations modulating the activity of the local circuit \cite{Okun:2012}.

@@

\subsection*{Combining pairwise interactions and coactivations}
In the effort to account for the joint activity patterns that are poorly explained by pairwise interactions, investigators have augmented models of pairwise interactions with additional mechanisms such as latent variables \cite{Koster:2013},  high-order correlations \cite{Ganmor:2011}, or global network fluctuations \cite{Tkacik:2013}.

@@

We showed that among several models a sparse network of linear interactions with several latent inputs yielded the best estimates of the noise covariance matrix for cortical microcircuits.  This finding is valuable in itself: improved estimates of the noise covariance matrix for large datasets are important in order to understand the role of noise correlations in population coding \cite{Abbott:1999, Sompolinsky:2001, Averbeck:2006, Ecker:2011}

@@

\subsection*{Surgery and two-photon imaging}
The surgical procedures and data acquisition were performed as described in \cite{Cotton:2013}: C57BL/6J mice (aged p40--60) were used. Anesthesia was initiated with isoflurane (3\%) and maintained with the mixture of fentanyl (0.05 mg/kg), midazolam (5 mg/kg), and medetomidine (0.5 mg/kg), with boosts of half the initial dose every 3 hours.  A craniotomy was performed over the right primary visual cortex.  Membrane-permeant calcium indicator Oregon Green 488 BAPTA-1 AM (OGB-1, Invitrogen) was loaded by bolus injection.  The craniotomy was sealed using a glass coverslip secured with dental cement.

@@

\subsection*{Visual stimulus}
The visual stimulus consisted of full-field drifting gratings with 90\% contrast, luminance of 10 cd/m$^2$, spatial frequency of 0.08 cycles/degree, and temporal frequency of 2 cycles/s. Two types of stimuli were presented for each imaging site: First, directional tuning was mapped using a pseudo-random sequence of drifting gratings at sixteen directions of motion, 500 ms per direction, without blanks, with 12--30 trials for each direction of motion.  Second, ro measure correlations, the stimulus was modified to include only two directions of motion (in 9 datasets) or five directions (in 22 datasets) and the gratings were presented for 1 second and were separated by 1-second blanks, with 100--300 trials for each direction of motion.

@@

In estimator $C_{\sf factor}$ (Eq.~\ref{eq:c-factor}), the low-rank matrix $L$ and the diagonal matrix $D$ are found by solving the minimization problem
\begin{equation}
(L,D) = \argmin\limits_{\hat L,\hat D} \loss{\hat L + \hat D,C_{\sf sample}},
\end{equation}
by an expectation-maximization (EM) algorithm with specified rank of $L$. Additional the diagonal matrix of individual variances is shrunk toward its mean value according to Eq.~\ref{eq:c-factor}.
